### 2014-03-21 Docs/architecture/architecture.tex
document multigamma + remove some old references to extern/

    The various problem-setups.  Each problem in \maestro\ gets it own
    sub-directory under {\tt SCIENCE/}, {\tt TEST\_PROBLEMS/}, or {\tt
      UNIT\_TESTS/}.  The {\tt GNUmakefile} in the problem directory
    includes the instructions on how to build the executable,
    including what modules in {\tt extern} are used.  Any file that
    you place in a sub-directory here takes precedence over a file of
    the same name in {\tt MAESTRO/}.  This allows problems to have
    custom versions of the main \maestro\ routines (e.g.\ initial
    conditions via {\tt initdata.f90}.  See \S~\ref{sec:makefile} and
    Chapter~\ref{ch:make} for details on the build system).

@@

Some packages (for instance, the {\tt helmeos}
EOS) require Fortran include files.  The {\tt Fmincludes} variable
lists all those directories that contain include files that are
inserted into the Fortran source at compile time via the {\tt include}
statement.  Presently, the only instance of this is with the Helmholtz
general equation of state found in {\tt extern/EOS/helmeos/}.  This is
automatically handled by the {\tt GMaestro.mak} instructions.

@@

\maestro\ initial models are read in by the {\tt extern/model\_parser}
routines.  This expects the initial model to contain a header giving
the number of variables and their names, followed by rows of data
giving the coordinate and data values at that coordinate.  The initial
model should contain the same species data (in the form of mass fractions) as
defined in the {\tt network} module used by the \maestro\ problem.

@@

Presently there are 2 equations of state that work with \maestro:
\begin{itemize}
\item {\tt extern/EOS/helmeos/} represents a general stellar equation 
      of state, consisting of nuclei (as an ideal gas), radiation,
      and electrons (with arbitrary degeneracy and degree of relativity).
      This equation of state is that described in \cite{timmes_eos}.

@@

\item {\tt extern/EOS/gamma\_law\_general} assumes an ideal gas with a mixed 
     composition and a constant ratio of specific heats, $\gamma$:
      \begin{equation}
      p = \rho e (\gamma - 1) = \frac{\rho k_B T}{\mu m_p} 
      \end{equation}
     where $k_B$ is Boltzmann's constant and $m_p$ is the mass of the
     proton.
     The mean molecular weight, $\mu$, is computed assuming 
     electrically neutral atoms:
     \begin{equation}
     \mu = \left ( \sum_k \frac{X_k}{A_k} \right )^{-1}
     \end{equation}
     An option in the source code itself exists for treating the
     species as fully-ionized, but there is no runtime-parameter to
     make this switch.
\end{itemize}

@@

The {\tt network} module defines the number species advected by the
code ({\tt nspec}), their ordering, and gives their basic properties
(like atomic number, $Z$, and atomic mass, $A$).  All \maestro\ problems
require a {\tt network} module, even if there are no reactions
modeled.  Many different reaction modules (containing different sets
of isotopes) exist in {\tt extern/networks}.  The particular network
used by a problem is defined in the problem's {\tt GNUmakefile}.

>>>>

    The various problem-setups.  Each problem in \maestro\ gets it own
    sub-directory under {\tt SCIENCE/}, {\tt TEST\_PROBLEMS/}, or {\tt
      UNIT\_TESTS/}.  The {\tt GNUmakefile} in the problem directory
    includes the instructions on how to build the executable,
    including what modules in {\tt Microphysics/} are used.  Any file that
    you place in a sub-directory here takes precedence over a file of
    the same name in {\tt MAESTRO/}.  This allows problems to have
    custom versions of the main \maestro\ routines (e.g.\ initial
    conditions via {\tt initdata.f90}.  See \S~\ref{sec:makefile} and
    Chapter~\ref{ch:make} for details on the build system).

@@

Some packages (for instance, the {\tt helmeos}
EOS) require Fortran include files.  The {\tt Fmincludes} variable
lists all those directories that contain include files that are
inserted into the Fortran source at compile time via the {\tt include}
statement.  Presently, the only instance of this is with the Helmholtz
general equation of state found in {\tt Microphysics/EOS/helmeos/}.  This is
automatically handled by the {\tt GMaestro.mak} instructions.

@@

\maestro\ initial models are read in by the {\tt Util/model\_parser}
routines.  This expects the initial model to contain a header giving
the number of variables and their names, followed by rows of data
giving the coordinate and data values at that coordinate.  The initial
model should contain the same species data (in the form of mass fractions) as
defined in the {\tt network} module used by the \maestro\ problem.

@@

Presently there are 3 equations of state that work with \maestro:
\begin{itemize}
\item {\tt helmeos/} represents a general stellar equation 
      of state, consisting of nuclei (as an ideal gas), radiation,
      and electrons (with arbitrary degeneracy and degree of relativity).
      This equation of state is that described in \cite{timmes_eos}.

@@

\item {\tt gamma\_law\_general} assumes an ideal gas with a mixed 
     composition and a constant ratio of specific heats, $\gamma$:
      \begin{equation}
      p = \rho e (\gamma - 1) = \frac{\rho k_B T}{\mu m_p} 
      \end{equation}
     where $k_B$ is Boltzmann's constant and $m_p$ is the mass of the
     proton.
     The mean molecular weight, $\mu$, is computed assuming 
     electrically neutral atoms:
     \begin{equation}
     \mu = \left ( \sum_k \frac{X_k}{A_k} \right )^{-1}
     \end{equation}
     An option in the source code itself exists for treating the
     species as fully-ionized, but there is no runtime-parameter to
     make this switch.

\item {\tt multigamma} is an ideal gas equation of state where each
  species can have a different value of $\gamma$.  This mainly affects
  how the internal energy is constructed as each species, represented
  with a mass fraction $X_k$ will have its contribution to the total
  specific internal energy take the form of $e = p/\rho/(\gamma_k -                                               
  1)$.  The main thermodynamic quantities take the form:
\begin{align}
p &= \frac{\rho k T}{m_u} \sum_k \frac{X_k}{A_k} \\
e &= \frac{k T}{m_u} \sum_k \frac{1}{\gamma_k - 1} \frac{X_k}{A_k} \\
h &= \frac{k T}{m_u} \sum_k \frac{\gamma_k}{\gamma_k - 1} \frac{X_k}{A_k}
\end{align}
We recognize that the usual astrophysical $\bar{A}^{-1} = \sum_k                                                  
X_k/A_k$, but now we have two other sums that involve different
$\gamma_k$ weightings.

The specific heats are constructed as usual,
\begin{align}
c_v &= \left . \frac{\partial e}{\partial T} \right |_\rho =
    \frac{k}{m_u} \sum_k \frac{1}{\gamma_k - 1} \frac{X_k}{A_k} \\
c_p &= \left . \frac{\partial h}{\partial T} \right |_p =
    \frac{k}{m_u} \sum_k \frac{\gamma_k}{\gamma_k - 1} \frac{X_k}{A_k}
\end{align}
and it can be seen that the specific gas constant, $R \equiv c_p - c_v$ is
independent of the $\gamma_i$, and is simply $R = k/m_u\bar{A}$ giving the
usual relation that $p = R\rho T$.  Furthermore, we can show
\begin{equation}
\Gamma_1 \equiv \left . \frac{\partial \log p}{\partial \log \rho} \right |_s =
   \left ( \sum_k \frac{\gamma_k}{\gamma_k - 1} \frac{X_k}{A_k} \right ) \bigg /
   \left ( \sum_k \frac{1}{\gamma_k - 1} \frac{X_k}{A_k} \right ) =
\frac{c_p}{c_v} \equiv \gamma_\mathrm{effective}
\end{equation}
and $p = \rho e (\gamma_\mathrm{effective} - 1)$.

This equation of state takes several runtime parameters that can set the
$\gamma_i$ for a specific species:
\begin{itemize}
\item {\tt eos\_gamma\_default}: the default $\gamma$ to apply for
  all species
\item {\tt species\_X\_name} and {\tt species\_X\_gamma}: set the $\gamma_i$
  for the species whose name is given as {\tt species\_X\_name} to the
  value provided by {\tt species\_X\_gamma}.  Here, {\tt X} can be one
  of the letters: {\tt a}, {\tt b}, or {\tt c}, allowing us to specify
  custom $\gamma_i$ for up to three different species.
\end{itemize}

@@

The {\tt network} module defines the number species advected by the
code ({\tt nspec}), their ordering, and gives their basic properties
(like atomic number, $Z$, and atomic mass, $A$).  All \maestro\ problems
require a {\tt network} module, even if there are no reactions
modeled.  Many different reaction modules (containing different sets
of isotopes) exist in {\tt Microphysics/networks}.  The particular network
used by a problem is defined in the problem's {\tt GNUmakefile}.
